---
title: "Multiple Regression Analysis"
author: "Autumn Pereira"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Prep

Let's read in our data and load some packages

```{r}
library(MASS)
library(ggplot2)
library(car)

d <- read.csv("complete_combined_spruce_data.csv")
```

And remove the ambient temperature treatments (based on observations from the exploratory data analysis) and omit any rows with NAs

```{r}
d <- subset(d, Temp_experimental != "Amb")
d <- na.omit(d)
```

Let's also log-transform MBC, MBN, bacteria copy number and archaea copy number

```{r}
d$Bacteria_copy_dry <- log(d$Bacteria_copy_dry)
d$Archaea_copy_dry <- log(d$Archaea_copy_dry)

d$MBC <- log(d$MBC+1)
d$MBN <- log(d$MBN+1)
```


We will make 2 smaller data frames containing only our predictors and response variable

```{r}
CO2 <- d[, c(4, 7:14)]
temp <- d[, c(3,7:14)]
```

And we will code the CO2 variable as binary, where 1= elevated CO2 and 0 = ambient CO2

```{r}
CO2$CO2 <- NA

for (i in 1:177){
  if (CO2$CO2_treatment[i] == "E"){
    CO2$CO2[i] = 1 }
  else{
   CO2$CO2[i] = 0 
  }
}

CO2 <- CO2[, -c(1)]
```

## Multicollinearity Concerns

We want to be sure that there is no multicollinearity in our data. From the correlation plots in the SEM file, it seems that all the correlations are < 0.6, so we will take these as not multicollinear enough for us to have a problem.

## Modelling Experimental CO2

We will fit a binomial GLM to the experimental CO2, and use stepwise AIC in both directions to select the most appropriate model.

```{r}
CO2_model <- glm(CO2 ~.,family=binomial(link='logit'),data=CO2)

CO2_step_model <- stepAIC(CO2_model, direction = "backward")
summary(CO2_step_model)

final_CO2_model <- glm(formula = CO2 ~ Bacteria_copy_dry + GWC + temp, family = binomial(link = "logit"), data = CO2)
```

Great, so the best model allowing us to parse out our ambient and elevated CO2 treatments includes

-Log bacterial copy number
- Gravimetric water content
- Temperature

The assumptions of logistic regression that we haven't checked is that there is a linear relationship between logit(CO2) and our response variables. We can test this using a Box-Tidwell test

```{r}
CO2$logits <- final_CO2_model$linear.predictors
```

```{r}
plot(CO2$logits, CO2$Bacteria_copy_dry)
plot(CO2$logits, CO2$GWC)
plot(CO2$logits, CO2$temp)
```

Although there is a lot of spread, there is prevalence of a linear relationship from these plots, so the model is valid.

## Modelling Experimental Temperature

Let's do the same for temperature. Here we will fit just a linear model since the temperatures are distributed linearly.

From the plots in the SEM RSheet, it seems that the independent variables are more or less linearly related to experimental temperature, so that satisfies that concern.

```{r}
temp_model <- lm(Temp_experimental ~., data = temp)
# Stepwise regression model
temp_step_model <- stepAIC(temp_model, direction = "backward", 
                      trace = TRUE)
summary(temp_step_model)
```

Great! So the best model in this case includes

- log bacteria copy number
- archaea copy number
- gravimetric water content
- dissolved nitrogen
- microbial biomass nitrogen
- dissolved organic carbon
- temperature

Let's look at the model assumptions

```{r}
plot(lm(formula = Temp_experimental ~ Bacteria_copy_dry + Archaea_copy_dry + 
    GWC + DN_unfumigated_soil + MBN + DOC_unfumigated_soil + 
    temp, data = temp))
```
The residuals vs fitted do look a bit odd, possibly because while our response variable takes values linearly distributed betweeen [0,9], it is categorical. Otherwise, the plots look acceptable. Otherwise, this follows the model assumptions quite well.
